A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-16 tidyindex.qmd
work on the GGGI example

The Global Gender Gap Index (GGGI), published annually by the World Economic Forum, measures gender parity by assessing relative gaps between men and women in four key areas: Economic Participation and Opportunity, Educational Attainment, Health and Survival, and Political Empowerment [@WEF2023]. The index is composed of 14 variables, expressed as female-to-male ratios, which are first aggregated in a linear combination into the four dimensions using the weight from the `V-weight` column in @tbl-gggi-weights. The weight is calculated as the inverse of the standard deviation of each variable and scaling to sum to 1 within each dimension to allow a one percentage point change in the standard deviation of each variable to contribute equally to the index. The four dimensions are then aggregated in another linear combination with equal weight to obtain the index. The 2023 GGGI data is available from the Global Gender Gap Report 2023 in the country's economy profile and can be accessed in R via the `tidyindex` package as `gggi`, along with the corresponding weights `gggi_weights`.

```{r tbl-gggi-weights}
#| tbl-cap: Weights of the fourteen variables in Global Gender Gap Index
#| tbl-cap-location: bottom
gggi_weights %>%  
   mutate(variable = str_replace_all(variable, "_", " ") %>%
            str_to_sentence()) %>% 
  select(c(1,2, 5:7)) %>% 
  mutate(dimension = c(rep("Economy", 5), rep("Education" ,4), 
                       rep("Health", 2), rep("Politics", 3))) %>% 
  knitr::kable(
    digits = 3, 
    col.names = c("Variable", "Dimension", "V-weight",
        "D-weight", "Weight")
    ) 
```

(Moved from the pipeline module to here) When a physical process is absent to describe the phenomenon, multivariate data are often combined using statistical methods, which can be further classified into linear and non-linear techniques. Many social indexes use a linear combination in an additive structure, $y(s; t) = \sum_{i = 1}^{p}\lambda_{i}x_i(s;t),$ where $\lambda_i$ is the weight assigned to each individual variable $x_i$. A linear combination can also be preceived as a linear projection of multivariate information through the weight vector. Projecting data from higher to lower dimension unavoidably leads to information loss and the used weights need careful examination to understand their effects on the index and implications for interpretation and decision-making. One can visualise the data projection across different sets of weights through animation using a method called tour.

A natural thing to do when provided with the index data is to reproduce the index. This helps index analysts to verify the index calculation and become familiar with the methodology. For GGGI, the construction can be simplified as a single linear aggregation step in the dimension reduction module, with the `Weight` column in @tbl-gggi-weights, which is the product of the variable weight (`V-weight`) and dimension weight (`D-weight`).

```{r fig-pipeline-steps-gggi}
#| fig-align: center
#| fig-cap: "Diagram of pipeline steps for index construction. will need to be updated with better design and the distribution fitting step."
#| out-height: 90%
#| out-width: 100%
knitr::include_graphics(here::here("figures/pipeline-gggi.png"))
```

    gggi %>% 
      init(id = country) %>%
      add_meta(gggi_weights, var_col = variable) %>% 
      dimension_reduction(
        index_new = aggregate_linear(
          ~labour_force_participation:years_with_female_head_of_state,
          weight = weight))

The result can be compared with the GGGI values available in the report, validating the reproducibility of the index for country with no missing variables.

```{r eval = FALSE}
a <- dim_weights %>% select(-variable) %>% t() %>% as_tibble(rownames = "id")
colnames(a) <- c("id", dim_weights$variable)
a %>% 
  pivot_longer(-id, names_to = "vars", values_to = "value") %>% 
  ggplot(aes(x = value, group = vars)) + 
  geom_histogram() + 
  facet_wrap(vars(vars))
```

```{r fig-compare-gggi}
#| fig-cap: "Verifying the calculation of the Global Gender Gap Index (GGGI). The index can be reproduced from the methodology described in the Global Gender Gap Report 2023 after removing the countries with missing variables, the treatment of which is unclear."
set.seed(123)
n <- 1000
w_idx <- map(1:n, function(x){
  w_idx <- stats::runif(4, min = 0.1, max = 1)
  w_idx <- w_idx/sum(w_idx)
  w_idx %>% as.matrix(nrow = 4, ncol = 1)
  })
rand_w <- reduce(w_idx, cbind) %>% as_tibble()
colnames(rand_w) <- paste0("V", 1:n)

dt2 <- gggi %>% 
  init(id = country) %>%
  add_meta(gggi_weights, var_col = variable) %>% 
  dimension_reduction(
    index_new = aggregate_linear(
      ~labour_force_participation:years_with_female_head_of_state,
      weight = weight)) 
```

```{r eval = FALSE, gggi-weights, fig.height=20, fig.align='center', fig.width=10}
dt3 <- dt2 %>%
  swap_values(.id = 1, .param = weight,
              .value = list(pillar_weight, paste0("V", 2:100)),
              .raw_data = dt)

idx_wide <- dt3$data %>% select(country, dplyr::contains("index_new"))
idx_wide2 <- idx_wide %>% mutate(index_avg = rowMeans(idx_wide[,2:(ncol(idx_wide))]))
idx_long2 <- idx_wide2 %>% pivot_longer(-country, names_to = "index", values_to = "value") %>% 
  left_join(gggi %>% select(country, rank), by = "country")

idx_long2 %>%
  ggplot(aes(x = value, y = fct_reorder(country, -rank), group = country)) +
  geom_boxplot() + 
  #geom_point(color = "grey80", size = 0.5) +
  geom_point(data = idx_long2 %>% filter(index == "index_new1"), color = "red") + 
  theme_bw() + 
  ylab("Country") + 
  xlab("Index")
```

>>>>

The Global Gender Gap Index (GGGI), published annually by the World Economic Forum, measures gender parity by assessing relative gaps between men and women in four key areas: Economic Participation and Opportunity, Educational Attainment, Health and Survival, and Political Empowerment [@WEF2023]. The index, compiled from 14 variables expressed as female-to-male ratios, first aggregates these variables into four dimensions through linear combinations. These dimensions are then combined though another linear combination. @fig-pp-gggi illustrates this pipeline construction by applying the dimension reduction module twice on the data to generate the index. @tbl-gggi-weights presents the The variable weights (`V-weight`) and dimension weights (`D-weight`) used in the two dimension reduction. In the table, the variable weights are computed as the inverse of the standard deviation of each variable, scaled to sum to 1. This weights ensures that a one percentage point change in the standard deviation of each variable to contribute equally to the index. Dimension weights are equal across the four dimensions and the last column, `weight`, multiples the variable and dimension weights to produce a single set of weight. The 2023 GGGI data is available from the Global Gender Gap Report 2023 in the country's economy profile and can be accessed in R via the `tidyindex` package as `gggi` and @tbl-gggi-weights as `gggi_weights`.

```{r fig-pp-gggi}
#| fig-align: center
#| fig-cap: "Index pipeline for the Global Gender Gap Index (GGGI). The index is constructed as applying the module dimension reduction twice on the data."
#| out-height: 90%
#| out-width: 100%
knitr::include_graphics(here::here("figures/pipeline-gggi.png"))
```

```{r tbl-gggi-weights}
#| tbl-cap: Weights used to compute the Global Gender Gap Index
#| tbl-cap-location: bottom
gggi_weights %>%  
   mutate(variable = str_replace_all(variable, "_", " ") %>%
            str_to_sentence()) %>% 
  select(c(1,2, 5:7)) %>% 
  mutate(dimension = c(rep("Economy", 5), rep("Education" ,4), 
                       rep("Health", 2), rep("Politics", 3))) %>% 
  knitr::kable(
    digits = 3, 
    col.names = c("Variable", "Dimension", "V-weight",
        "D-weight", "Weight")
    ) 
```

The index can be reproduce with tidyindex as:

```{r eval = FALSE, echo = TRUE}
gggi %>% 
  init(id = country) %>%
  add_meta(gggi_weights, var_col = variable) %>% 
  dimension_reduction(
    index_new = aggregate_linear(
      ~labour_force_participation:years_with_female_head_of_state,
      weight = weight)) 
```

After initialising the `gggi` object and attaching the `gggi_weights` as metadata, a single linear aggregation step within the dimension reduction module is applied to the 14 variables (from column `labour_force_participation` to `years_with_female_head_of_state`), using the weight specified in the `weight` column of the attached metadata. While computing the index from the original 14 variables, it remains unclear how the missing values are handled within the index, which impacts 68 out of the total 146 countries. However, after aggregating variables into the four dimensions, where no missing values exist, the index is reproducible for all the countries.

A linear combination can also be perceived as a linear projection of multivariate information through the weight vector. Projecting data from higher to lower dimension unavoidably leads to information loss and the used weights need careful examination to understand their effects on the index and implications for interpretation and decision-making. One can visualise the data projection across different sets of weights through animation using a method called tour.

### 2023-08-16 tidyindex.tex
work on the GGGI example

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, breakable, interior hidden]}{\end{tcolorbox}}\fi

@@

The Global Gender Gap Index (GGGI), published annually by the World
Economic Forum, measures gender parity by assessing relative gaps
between men and women in four key areas: Economic Participation and
Opportunity, Educational Attainment, Health and Survival, and Political
Empowerment (World Economic Forum 2023). The index is composed of 14
variables, expressed as female-to-male ratios, which are first
aggregated in a linear combination into the four dimensions using the
weight from the \texttt{V-weight} column in
Table~\ref{tbl-gggi-weights}. The weight is calculated as the inverse of
the standard deviation of each variable and scaling to sum to 1 within
each dimension to allow a one percentage point change in the standard
deviation of each variable to contribute equally to the index. The four
dimensions are then aggregated in another linear combination with equal
weight to obtain the index. The 2023 GGGI data is available from the
Global Gender Gap Report 2023 in the country's economy profile and can
be accessed in R via the \texttt{tidyindex} package as \texttt{gggi},
along with the corresponding weights \texttt{gggi\_weights}.

\hypertarget{tbl-gggi-weights}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.5455}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1299}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1169}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1169}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0909}}@{}}
\caption{\label{tbl-gggi-weights}Weights of the fourteen variables in
Global Gender Gap Index}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Dimension
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
V-weight
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
D-weight
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Weight
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Dimension
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
V-weight
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
D-weight
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Weight
\end{minipage} \\
\midrule()
\endhead
Labour force participation & Economy & 0.199 & 0.25 & 0.050 \\
Wage equality for similar work & Economy & 0.310 & 0.25 & 0.078 \\
Estimated earned income & Economy & 0.221 & 0.25 & 0.055 \\
Legislators senior officials and managers & Economy & 0.149 & 0.25 &
0.037 \\
Professional and technical workers & Economy & 0.121 & 0.25 & 0.030 \\
Literacy rate & Education & 0.191 & 0.25 & 0.048 \\
Enrolment in primary education & Education & 0.459 & 0.25 & 0.115 \\
Enrolment in secondary education & Education & 0.230 & 0.25 & 0.058 \\
Enrolment in tertiary education & Education & 0.121 & 0.25 & 0.030 \\
Sex ratio at birth & Health & 0.693 & 0.25 & 0.173 \\
Healthy life expectancy & Health & 0.307 & 0.25 & 0.077 \\
Women in parliament & Politics & 0.310 & 0.25 & 0.078 \\
Women in ministerial positions & Politics & 0.247 & 0.25 & 0.062 \\
Years with female head of state & Politics & 0.443 & 0.25 & 0.111 \\
\bottomrule()
\end{longtable}

(Moved from the pipeline module to here) When a physical process is
absent to describe the phenomenon, multivariate data are often combined
using statistical methods, which can be further classified into linear
and non-linear techniques. Many social indexes use a linear combination
in an additive structure,
\(y(s; t) = \sum_{i = 1}^{p}\lambda_{i}x_i(s;t),\) where \(\lambda_i\)
is the weight assigned to each individual variable \(x_i\). A linear
combination can also be preceived as a linear projection of multivariate
information through the weight vector. Projecting data from higher to
lower dimension unavoidably leads to information loss and the used
weights need careful examination to understand their effects on the
index and implications for interpretation and decision-making. One can
visualise the data projection across different sets of weights through
animation using a method called tour.

A natural thing to do when provided with the index data is to reproduce
the index. This helps index analysts to verify the index calculation and
become familiar with the methodology. For GGGI, the construction can be
simplified as a single linear aggregation step in the dimension
reduction module, with the \texttt{Weight} column in
Table~\ref{tbl-gggi-weights}, which is the product of the variable
weight (\texttt{V-weight}) and dimension weight (\texttt{D-weight}).

@@

\caption{\label{fig-pipeline-steps-gggi}Diagram of pipeline steps for
index construction. will need to be updated with better design and the
distribution fitting step.}

@@

\begin{verbatim}
gggi %>% 
  init(id = country) %>%
  add_meta(gggi_weights, var_col = variable) %>% 
  dimension_reduction(
    index_new = aggregate_linear(
      ~labour_force_participation:years_with_female_head_of_state,
      weight = weight)) 
\end{verbatim}

The result can be compared with the GGGI values available in the report,
validating the reproducibility of the index for country with no missing
variables.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, enhanced, breakable]}{\end{tcolorbox}}\fi

@@

The Global Gender Gap Index (GGGI), published annually by the World
Economic Forum, measures gender parity by assessing relative gaps
between men and women in four key areas: Economic Participation and
Opportunity, Educational Attainment, Health and Survival, and Political
Empowerment (World Economic Forum 2023). The index, compiled from 14
variables expressed as female-to-male ratios, first aggregates these
variables into four dimensions through linear combinations. These
dimensions are then combined though another linear combination.
Figure~\ref{fig-pp-gggi} illustrates this pipeline construction by
applying the dimension reduction module twice on the data to generate
the index. Table~\ref{tbl-gggi-weights} presents the The variable
weights (\texttt{V-weight}) and dimension weights (\texttt{D-weight})
used in the two dimension reduction. In the table, the variable weights
are computed as the inverse of the standard deviation of each variable,
scaled to sum to 1. This weights ensures that a one percentage point
change in the standard deviation of each variable to contribute equally
to the index. Dimension weights are equal across the four dimensions and
the last column, \texttt{weight}, multiples the variable and dimension
weights to produce a single set of weight. The 2023 GGGI data is
available from the Global Gender Gap Report 2023 in the country's
economy profile and can be accessed in R via the \texttt{tidyindex}
package as \texttt{gggi} and Table~\ref{tbl-gggi-weights} as
\texttt{gggi\_weights}.

@@

\caption{\label{fig-pp-gggi}Index pipeline for the Global Gender Gap
Index (GGGI). The index is constructed as applying the module dimension
reduction twice on the data.}

@@

\hypertarget{tbl-gggi-weights}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.5455}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1299}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1169}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1169}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0909}}@{}}
\caption{\label{tbl-gggi-weights}Weights used to compute the Global
Gender Gap Index}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Dimension
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
V-weight
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
D-weight
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Weight
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Dimension
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
V-weight
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
D-weight
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Weight
\end{minipage} \\
\midrule()
\endhead
Labour force participation & Economy & 0.199 & 0.25 & 0.050 \\
Wage equality for similar work & Economy & 0.310 & 0.25 & 0.078 \\
Estimated earned income & Economy & 0.221 & 0.25 & 0.055 \\
Legislators senior officials and managers & Economy & 0.149 & 0.25 &
0.037 \\
Professional and technical workers & Economy & 0.121 & 0.25 & 0.030 \\
Literacy rate & Education & 0.191 & 0.25 & 0.048 \\
Enrolment in primary education & Education & 0.459 & 0.25 & 0.115 \\
Enrolment in secondary education & Education & 0.230 & 0.25 & 0.058 \\
Enrolment in tertiary education & Education & 0.121 & 0.25 & 0.030 \\
Sex ratio at birth & Health & 0.693 & 0.25 & 0.173 \\
Healthy life expectancy & Health & 0.307 & 0.25 & 0.077 \\
Women in parliament & Politics & 0.310 & 0.25 & 0.078 \\
Women in ministerial positions & Politics & 0.247 & 0.25 & 0.062 \\
Years with female head of state & Politics & 0.443 & 0.25 & 0.111 \\
\bottomrule()
\end{longtable}

The index can be reproduce with tidyindex as:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{gggi }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{init}\NormalTok{(}\AttributeTok{id =}\NormalTok{ country) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{add\_meta}\NormalTok{(gggi\_weights, }\AttributeTok{var\_col =}\NormalTok{ variable) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{dimension\_reduction}\NormalTok{(}
    \AttributeTok{index\_new =} \FunctionTok{aggregate\_linear}\NormalTok{(}
      \SpecialCharTok{\textasciitilde{}}\NormalTok{labour\_force\_participation}\SpecialCharTok{:}\NormalTok{years\_with\_female\_head\_of\_state,}
      \AttributeTok{weight =}\NormalTok{ weight)) }
\end{Highlighting}
\end{Shaded}

After initialising the \texttt{gggi} object and attaching the
\texttt{gggi\_weights} as metadata, a single linear aggregation step
within the dimension reduction module is applied to the 14 variables
(from column \texttt{labour\_force\_participation} to
\texttt{years\_with\_female\_head\_of\_state}), using the weight
specified in the \texttt{weight} column of the attached metadata. While
computing the index from the original 14 variables, it remains unclear
how the missing values are handled within the index, which impacts 68
out of the total 146 countries. However, after aggregating variables
into the four dimensions, when no missing values exist, the index is
reproducible for all the countries.

A linear combination can also be perceived as a linear projection of
multivariate information through the weight vector. Projecting data from
higher to lower dimension unavoidably leads to information loss and the
used weights need careful examination to understand their effects on the
index and implications for interpretation and decision-making. One can
visualise the data projection across different sets of weights through
animation using a method called tour.
